Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Re-written text about overall slowdowns/speedups. Removed one plot as essentially duplicated the plot of perf. of original code

In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the Shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and green bars. The performance of the Ivy
Bridge system with the Cray or Intel compiler (red and purple bars)
only drops to this level when the domain size is increased to
$1024^{2}$. At this point, the working set no longer fits within cache
and performance is dominated by the bandwidth to main memory, making
it relatively insensitive to the choice of compiler.

@@

For this compiler-friendly form of the code, there is little
performance difference between the executables produced by the Gnu and
Intel compilers. However, the executable produced by the Cray compiler
generally performs significantly better for all but the largest domain
size.

Moving now the to the \psykal version of Shallow,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal version for each of the compiler/CPU combinations. The
only significant difference between this and the plot of the
performance of the original version of the code in
Figure~\ref{FIG_orig_perf_summary} is that the difference between the
Gnu- and Intel-compiled executables is slightly more marked. This is
clarified in Figure~\ref{FIG_slowdown_summary} which plots the
percentage difference between the performance of the original and the
(best) \psykal versions of Shallow for each compiler/CPU combination.
It is clear from this plot that the difference in the performance of
the Gnu- and Intel-compiled excutables on the Haswell system is due to
the Gnu compiler being unable to recover the performance achieved by
the original code. In contrast there is no such problem for the
Gnu-compiled executable on the Ivy Bridge system. We attribute this to
the more recent version of the compiler on the latter system (4.9.1 as
opposed to 4.8.2).

\begin{figure}[!t]
\centering
\includegraphics[width=120mm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal 
version of Shallow with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

@@

In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
\psykal version of Shallow is never more than 2\% slower than the
original and in some cases is faster.  This demonstrates that we can
reliably recover the performance of the original version of the code,
despite the significant restructuring required by the \psykal
approach.

The most striking feature of Figure~\ref{FIG_slowdown_summary} is the
fact that the best \psykal version of Shallow is consistently faster
than the original on the Power 8. This confirms that this early
version of the IBM compiler struggled with the original form of the
Shallow source code.

>>>>

In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the Shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the first two bars. The performance of the Ivy
Bridge system with the Cray or Intel compiler (penultimate two bars)
only drops to this level when the domain size is increased to
$1024^{2}$. At this point, the working set no longer fits within cache
and performance is dominated by the bandwidth to main memory, making
it relatively insensitive to the choice of compiler.

@@

For this compiler-friendly form of the code, there is generally little
performance difference between the executables produced by the Gnu and
Intel compilers on both the Haswell and Ivy Bridge CPUs. However, the
executable produced by the Cray compiler generally performs
significantly better for all but the largest domain size. Analysis of
the optimisation reports produced by the Cray and Intel compilers did
not give any clues as to the cause of this difference and an in-depth
investigation is outside the scope of this paper.

Moving now the to the \psykal version of Shallow, it turns out that a
plot of the performance of the best performing \psykal version for
each of the compiler/CPU combinations is very similar to
Figure~\ref{FIG_orig_perf_summary}. Therefore, in
Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
  difference} between the performance of the original and the (best)
\psykal versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal approach.

The exceptional case is that of the Gnu-compiled binary on the Haswell
CPU for a problem size of $256^2$.  This is strange since the
performance difference for the same case on the Ivy Bridge CPU is just
0.8\%. {\bf installing 4.9.1 on desktop so can try running this case
  with it. In fact there are considerable difference between
  slow-downs seen with 4.8.2 and 4.9.3 on desktop -things are much
  better with more recent compiler.}

@@

Turning to the cases where the \psykal version is significantly
{\em faster} than the original, we see from
Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
this only occurs for the smallest domain size. This is no doubt due in
part to the fact that the problem executes very quickly and therefore
a small reduction in run-time will show as a large precentage. Since
this problem is well contained in L2 cache, its performance will also
be more sensitive to small differences, {\it e.g.} in the way that CPU
registers are used.

The performance of the Power 8 system with the $512^2$ domain size is
somewhat anomolous: the \psykal version of the code performs
significantly better than the original and in addition, this problem
size is also the only one where the Power 8 out-performs the Haswell
system (see Figure~\ref{FIG_orig_perf_summary}) for the original
version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
that the $512^2$ case marks a transition from the working set being in
cache (sizes $\leq 256$) to spilling out of cache (size 1024). We
therefore suggest that cache is being used more efficiently on the
Power 8 system and this is further enhanced by the IBM-supplied
transformations in the \psykal version.
